Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-02-27 analysis/paper/paper.Rmd
change default knit format to HTML
modify paper to frontiersin guidelines

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  fig.align = "center")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
# cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
# cite_keys <- unlist(cite_keys, use.names = FALSE)
# rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
```

@@

<!-- insert figure of the setup -->

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
```

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.

@@

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the system.

@@

```{r sample-perc}
prop_wide <- prop %>%
  ungroup() %>%
  pivot_wider(id_cols = c(treatment, starch), 
              names_from = size, 
              values_from = prop) %>%
  add_row(prop_comb_row, .after = 2)

perc_tbl <- prop %>%
  ungroup() %>%
  select(!c(sample, solution, prop)) %>%
  pivot_wider(names_from = size, values_from = perc) %>%
  add_row(perc_comb_row, .after = 2)
```

@@

```{r}
sol_size_prop <- sol_comb_out %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "solution")

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

<!--
```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```
-->

@@

<!-- limitations of the platform show which questions would be unanswered -->
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
the absence can also allow us to directly explore
the effect of $\alpha$-amylase on starch counts in future experiments,
where $\alpha$-amylase can be added to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

@@

@@

# Data Availability Statement {-}

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
following the format provided by the rrtools package [@R-rrtools].
More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/workspaces/byoc).
Additional tables and figures are available in the Supplementary Material.

# References cited

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
#author: |
# Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
# date: |
#  $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
#### Frontiers submission ####
runtitle: "Biases Associated with Starch Incorporation"
documentclass: frontiers/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
author:
  - name: Bjørn Peare Bartholdy
    affiliation: '1'
    etal: BARTHOLDY # First author's last name.
    email: b.p.bartholdy@arch.leidenuniv.nl # Indicates corresponding Author
    institution: Leiden University
    street: Einsteinweg 2
    city: Leiden
    zip: 2333 CC
    country: Netherlands
  - name: Amanda G. Henry
    affiliation: '1'
affiliation:
  - id: '1'
    department: Archaeological Sciences
    institution: Leiden University
    city: Leiden
    country: Netherlands
#### End: Frontiers submission ####
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: frontiers/frontiers.csl
params:
  keywords: 
    - oral biofilm model
    - starch
    - alpha-amylase
    - dental calculus
    - paleodiet reconstruction
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  #fig.path = "frontiers/figures/",
  fig.path = "../figures/",
  fig.align = "center",
  #fig.ext = "jpg",
  fig.pos = "H",
  dpi = 300) 
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor2_r <- signif(count_cor$estimate, 3)
cor2_ci <- signif(count_cor$conf.int, 3)
```

\begin{abstract}

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

\tiny
 \keyFont{ \section{Keywords:} `r params$keywords`}

\end{abstract}

@@

@@

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.

@@

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
(saliva)
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the model.

@@

```{r sample-perc}
prop_wide <- prop %>%
  ungroup() %>%
  pivot_wider(id_cols = c(treatment, starch), 
              names_from = size, 
              values_from = prop) %>%
  add_row(prop_comb_row, .after = 2)
```

@@

```{r size-trends}
sol_size_prop <- sol_comb_out %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "solution")

@@

@@

@@

<!-- limitations of the platform show which questions would be unanswered -->
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
also affect biofilm growth due to the lack of amylase-bacterium interactions
[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
absence of $\alpha$-amylase, because it can allow us to directly explore
its effect on starch counts in future experiments, where $\alpha$-amylase can be 
added to the model in concentrations similar to those found in the oral cavity
[@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

@@

# Open Science Statement {-}

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
following the format provided by the rrtools package [@R-rrtools].
More detailed protocols are available on OSF (https://osf.io/akevs/)
and protocols.io (https://www.protocols.io/workspaces/byoc).
Additional tables and figures are available in the Supplementary Material
(https://osf.io/ucxsv/).

# Author Contributions {-}

AGH acquired the funding for the analysis. BPB and AGH conceptualised and designed
the study. BPB conducted the experiments and analysed the data.
BPB wrote the manuscript. AGH reviewed and edited the manuscript.

# Funding {-}

# Conflict of Interest {-}

The authors declare that the research was conducted in the absence of any
commercial or financial relationships that could be construed as a potential
conflict of interest.

@@

# References {-}

::: {#refs}
:::

\clearpage

# Tables

# Figure Captions

```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup.", out.width="70%"}
knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
#ggsave("frontiers/figures/cor-plot.png", dpi = 300)
```

<!--
Figure 1: Overview of experiment protocol including the plate setup.

Figure 2: Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.

Figure 3: Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.

Figure 4: Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.
-->
